---
title: "Global climate change and contributing factors - Data Exploration"
author: "Merwa Tariq"
format: html
server: shiny
editor_options: 
  chunk_output_type: inline
---

## Shiny Documents

This Quarto document is made interactive using Shiny. Interactive documents allow readers to modify parameters and see the results immediately. Learn more about Shiny interactive documents at <https://quarto.org/docs/interactive/shiny/>.

## Inputs and Outputs

You can embed Shiny inputs and outputs in your document. Outputs are automatically updated whenever inputs change. This demonstrates how a standard R plot can be made interactive:

```{r}
library(shiny)
library(tidyverse)
library(dplyr)
library(maps)
library(ggrepel)
library(readxl)
library(scales)
library(shiny)
```

#world map data 
```{r}
world <- map_data("world")
```




#temperature data cleaning and exploration

```{r}
temperature <- read_csv('data/FAOSTAT_data.csv', show_col_types=FALSE) |>
  filter(Months=='Meteorological year') |>
  select(Area, Year, Unit, Value ) 
#181 x 64
```


```{r}
#for data cleaning
anti_join(world, temperature, 
          join_by(region==Area)) |>
  select(region) |>
  arrange(region) |>
  unique()
```

```{r}
#for data cleaning
anti_join(temperature,world , 
          join_by(Area==region)) |>
  select(Area) |>
  arrange(Area) |>
  unique() |>
  print(n=56)
```


```{r}
temperature <- temperature |>
  mutate(Area = case_when(
  	Area=="Bolivia (Plurinational State of)" ~ "Bolivia",
  	Area=="Brunei Darussalam" ~ "Brunei",
    Area=="Cabo Verde" ~ "Cape Verde",
  	Area=="China, Hong Kong SAR" ~ "Hong Kong",
    Area=="China, Macao SAR" ~ "Macao",
  	Area=="China, Taiwan Province of" ~ "Taiwan",
    Area=="China, mainland" ~ "NA",
    Area=="Cocos (Keeling) Islands" ~ "Cocos Islands",
  	Area=="Congo" ~ "Republic of the Congo",
  	Area=="Czechia" ~ "Czech Republic",
  	Area=="Democratic People's Republic of Korea" ~ "North Korea",
  	Area=="Eswatini" ~ "Swaziland",
  	Area=="Ethiopia PDR" ~ "Ethiopia",
  	Area=="Falkland Islands (Malvinas)" ~ "Falkland Islands",
  	Area=="French Guyana" ~ "French Guiana",
  	Area=="French Southern Territories" ~ "French Southern and Antarctic Lands",
  	Area=="Iran (Islamic Republic of)" ~ "Iran",
  	Area=="Lao People's Democratic Republic" ~ "Laos",
  	Area=="Micronesia (Federated States of)" ~ "Micronesia",
  	Area=="Pitcairn" ~ "Pitcairn Islands",
  	Area=="Republic of Korea" ~ "South Korea",
  	Area=="Republic of Moldova" ~ "Moldova",
  	Area=="Russian Federation" ~ "Russia",
  	Area=="Saint Helena, Ascension and Tristan da Cunha" ~ "Saint Helena",
  	Area=="Saint Kitts and Nevis" ~ "Saint Kitts",
  	Area=="Saint Kitts and Nevis" ~ "Nevis",
    Area=="Saint Vincent and the Grenadines" ~ "Saint Vincent",
    Area=="Saint Vincent and the Grenadines" ~ "Grenadines",
  	Area=="South Georgia and the South Sandwich Islands" ~ "South Georgia",
  	Area=="South Georgia and the South Sandwich Islands" ~ "South Sandwich Islands",
  	Area=="Syrian Arab Republic" ~ "Syria",
  	Area=="Trinidad and Tobago" ~ "Tobago",
  	Area=="Trinidad and Tobago" ~ "Trinidad",
    Area=="United Kingdom of Great Britain and Northern Ireland" ~ "UK",
    Area=="United Republic of Tanzania" ~ "Tanzania",
    Area=="United States of America" ~ "USA",
    Area=="Venezuela (Bolivarian Republic of)" ~ "Venezuela",
    Area=="Viet Nam" ~ "Vietnam",
    Area=="United States Virgin Islands" ~ "Virgin Islands",
    Area=="Wallis and Futuna Islands" ~ "Wallis and Futuna",
    TRUE ~ Area)) 
```


```{r}
temperature_new|>
  select (Area) |>
  group_by(Area) |>
  count() |>
  ggplot() +
  aes(x=n) +
  geom_histogram(bins=10) +
  theme_minimal() +
  labs(x="Number of records",
       title="Number of records present for Countries") +
  theme(axis.title.y = element_blank())

temperature_new|>
  select (Area) |>
  group_by(Area) |>
  count() |>
  arrange(n) |>
  filter(n < 60 | n>60) |>
  print(n=80)
```


```{r}
rep_2 <- temperature |>
  filter(Area %in% c("Antigua and Barbuda", "Belgium-Luxembourg", "Czechoslovakia","Saint Kitts and Nevis", "Saint Vincent and the Grenadines", "Serbia and Montenegro", "South Georgia and the South Sandwich Islands", "Sudan (former)", "Trinidad and Tobago"))

temperature_new <- rbind(rep_2 |> 
        mutate(Area = case_when(
          Area=="Antigua and Barbuda" ~ "Antigua",
          Area=="Belgium-Luxembourg" & Year<2000  ~ "Belgium",
          Area=="Czechoslovakia" & Year<1993  ~ "Czech Republic",
        	Area=="Saint Kitts and Nevis" ~ "Nevis",
          Area=="Saint Vincent and the Grenadines" ~ "Grenadines",
          Area=="Serbia and Montenegro" & Year>=1992 & Year<2006 ~ "Montenegro",
        	Area=="South Georgia and the South Sandwich Islands" ~ "South Georgia",
        	Area=="Trinidad and Tobago" ~ "Tobago",
        	Area=="Sudan (former)" & Year<2011 ~ "South Sudan",
          TRUE ~ Area)
      ), 
      rep_2 |> 
        mutate(Area = case_when(
          Area=="Antigua and Barbuda" ~ "Barbuda",
          Area=="Belgium-Luxembourg"  & Year<2000 ~ "Luxembourg",
          Area=="Czechoslovakia" & Year<1993  ~ "Slovakia",
        	Area=="Saint Kitts and Nevis" ~ "Saint Kitts",
          Area=="Saint Vincent and the Grenadines" ~ "Saint Vincent",
          Area=="Serbia and Montenegro" & Year>=1992 & Year<2006 ~ "Serbia",
        	Area=="South Georgia and the South Sandwich Islands" ~ "South Sandwich Islands",
        	Area=="Trinidad and Tobago" ~ "Trinidad",
        	Area=="Sudan (former)" & Year<2011 ~ "Sudan",
          TRUE ~ Area)
      ),
      temperature |>
        filter(!Area %in% c("Antigua and Barbuda", "Belgium-Luxembourg", "Czechoslovakia","Saint Kitts and Nevis", "Saint Vincent and the Grenadines", "Serbia and Montenegro" ,"South Georgia and the South Sandwich Islands", "Sudan (former)", "Trinidad and Tobago"))
)
```

```{r}
rep_5 <- temperature_new |>
  filter(Area=="Netherlands Antilles (former)")

temperature_new <- rbind(rep_5 |> 
        mutate(Area = case_when(
          Area=="Netherlands Antilles (former)" ~ "Curacao",
          TRUE ~ Area)
      ), 
      rep_5 |> 
        mutate(Area = case_when(
          Area=="Netherlands Antilles (former)" ~ "Sint Maarten",
          TRUE ~ Area)
      ),
      rep_5 |> 
        mutate(Area = case_when(
          Area=="Netherlands Antilles (former)" ~ "Bonaire",
          TRUE ~ Area)
      ),
      rep_5 |> 
        mutate(Area = case_when(
          Area=="Netherlands Antilles (former)" ~ "Saba",
          TRUE ~ Area)
      ),
      rep_5 |> 
        mutate(Area = case_when(
          Area=="Netherlands Antilles (former)" ~ "Sint Eustatius",
          TRUE ~ Area)
      ),
      temperature_new |>
        filter(Area!="Netherlands Antilles (former)")
)
```


```{r}
rep_USSR <- temperature_new |>
  filter(Area %in% c("USSR"))

temperature_new <- rbind(rep_USSR |> 
        mutate(Area = case_when(
          Area=="USSR" & Year<1992 ~ "Russia",
          TRUE ~ Area)
      ), 
      rep_USSR |> 
        mutate(Area = case_when(
          Area=="USSR" & Year<1992 ~ "Ukraine",
          TRUE ~ Area)
      ),
      rep_USSR |> 
        mutate(Area = case_when(
          Area=="USSR" & Year<1992 ~ "Belarus",
          TRUE ~ Area)
      ),
      rep_USSR |> 
        mutate(Area = case_when(
          Area=="USSR" & Year<1992 ~ "Moldova",
          TRUE ~ Area)
      ),
      rep_USSR |> 
        mutate(Area = case_when(
          Area=="USSR" & Year<1992 ~ "Lithuania",
          TRUE ~ Area)
      ),
      rep_USSR |> 
        mutate(Area = case_when(
          Area=="USSR" & Year<1992 ~ "Latvia",
          TRUE ~ Area)
      ),
      rep_USSR |> 
        mutate(Area = case_when(
          Area=="USSR" & Year<1992 ~ "Estonia",
          TRUE ~ Area)
      ),
      rep_USSR |> 
        mutate(Area = case_when(
          Area=="USSR" & Year<1992 ~ "Georgia",
          TRUE ~ Area)
      ),
      rep_USSR |> 
        mutate(Area = case_when(
          Area=="USSR" & Year<1992 ~ "Armenia",
          TRUE ~ Area)
      ),
      rep_USSR |> 
        mutate(Area = case_when(
          Area=="USSR" & Year<1992 ~ "Azerbaijan",
          TRUE ~ Area)
      ),
      rep_USSR |> 
        mutate(Area = case_when(
          Area=="USSR" & Year<1992 ~ "Kazakhstan",
          TRUE ~ Area)
      ),
      rep_USSR |> 
        mutate(Area = case_when(
          Area=="USSR" & Year<1992 ~ "Uzbekistan",
          TRUE ~ Area)
      ),
      rep_USSR |> 
        mutate(Area = case_when(
          Area=="USSR" & Year<1992 ~ "Kyrgyzstan",
          TRUE ~ Area)
      ),
      rep_USSR |> 
        mutate(Area = case_when(
          Area=="USSR" & Year<1992 ~ "Tajikistan",
          TRUE ~ Area)
      ),
      rep_USSR |> 
        mutate(Area = case_when(
          Area=="USSR" & Year<1992 ~ "Turkmenistan",
          TRUE ~ Area)
      ),
      temperature_new |>
        filter(Area!="USSR")
)
```


```{r}
rep_Yugoslav <- temperature_new |>
  filter(Area == "Yugoslav SFR")

temperature_new <- rbind(rep_Yugoslav |> 
        mutate(Area = case_when(
          Area=="Yugoslav SFR" & Year<1992 ~ "Croatia",
          TRUE ~ Area)
      ),
      rep_Yugoslav |> 
        mutate(Area = case_when(
          Area=="Yugoslav SFR" & Year<1992 ~ "Montenegro",
          TRUE ~ Area)
      ),
      rep_Yugoslav |> 
        mutate(Area = case_when(
          Area=="Yugoslav SFR" & Year<1992 ~ "Serbia",
          TRUE ~ Area)
      ),
      rep_Yugoslav |> 
        mutate(Area = case_when(
          Area=="Yugoslav SFR" & Year<1992 ~ "Slovenia",
          TRUE ~ Area)
      ),
      rep_Yugoslav |> 
        mutate(Area = case_when(
          Area=="Yugoslav SFR" & Year<1992 ~ "Bosnia and Herzegovina",
          TRUE ~ Area)
      ),
      rep_Yugoslav |> 
        mutate(Area = case_when(
          Area=="Yugoslav SFR" & Year<1992 ~ "North Macedonia",
          TRUE ~ Area)
      ),
      temperature_new |>
        filter(Area != "Yugoslav SFR")
)
```



```{r}
temperature <- temperature_new
rm(temperature_new)
```


#temperature change trend
```{r}
temperature|>
  ggplot() +
  aes(x=Year, y=Value, colour=Area, group=Area ) +
  geom_line() +
  labs(x='Year',
       y='Temperature Change in *C',
       title = "Temperature change across countries over the years" ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), 
        legend.position = "none") +
  scale_x_continuous(breaks = seq(1960, 2020, by = 5), expand = c(0, NA))
```


```{r}
latest_temp_change <- temperature |>
  drop_na(Value, Year) |>
  group_by(Area) |>
  do(model = lm(Value ~ as.numeric(Year), data = .)) |>
  summarise(
    Country = first(Area),
    intercept = coef(model)[1],
    slope = coef(model)[2]
  ) |>
  mutate(temp_change = slope*(2022)+intercept) |>
  select(Country, temp_change)
```


#temperature change across the world
```{r}
inner_join(latest_temp_change, world, join_by(Country==region)) |>
  ggplot() +
  aes(x=long, y=lat, group=group) + 
  coord_fixed(1.3) +
  geom_polygon(aes(fill = temp_change)) +
  theme_minimal() +
  scale_fill_gradient(low = "yellow", high = "red") + 
  labs(title = "Global temperature change") +
  theme(axis.title=element_blank(), 
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.title = element_blank(),
        legend.position = "bottom",
        panel.grid = element_blank(),
        plot.title = element_text(hjust = 0.5))
```






















#co2 emission data cleaning and exploration


```{r}
co2_emission <- read_csv('data/CO2 emission by countries.csv', show_col_types=FALSE) 
```


```{r}
#for data cleaning
anti_join(world, co2_emission, 
          join_by(region==Country)) |>
  select(region) |>
  arrange(region) |>
  unique()

anti_join(co2_emission, world,
          join_by(Country==region)) |>
  select(Country) |>
  arrange(Country) |>
  unique()
```

```{r}
co2_emission <- co2_emission |>
  mutate(Country = case_when(
    Country=="Congo" ~ "Republic of Congo",
    Country=="Cote d'Ivoire" ~ "Ivory Coast",
    Country=="Czechia" ~ "Czech Republic",
    Country=="Democratic Republic of Congo" ~ "Democratic Republic of the Congo",
    Country=="Eswatini" ~ "Swaziland",
    Country=="Faeroe Islands" ~ "Faroe Islands",
    Country=="Sint Maarten (Dutch part)" ~ "Sint Maarten",
    Country=="Timor" ~ "Timor-Leste",
    Country=="United Kingdom" ~ "UK",
    Country=="United States" ~ "USA",
    .default = Country),
    ) |> 
  mutate(
    `CO2 emission (Tons)` = round(`CO2 emission (Tons)`/10**9),
    ) 
```


#to view overall trend
```{r}
co2_emission |>
  filter(Year>1960 & Year<2023) |>
  group_by(Year) |>
  summarise(sum=sum(`CO2 emission (Tons)`)) |> 
  ggplot() +
  aes(x=Year, y=sum) +
  geom_line() +
  labs(x='Year',
       y='CO2 Emission in Billion Tons',
       title = "Global CO2 Emission" ) +
  scale_x_continuous(breaks = seq(1965, 2020, by = 5), expand = c(0, NA)) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

#to find important countries
```{r}
co2_emission |>
  select(Country, `CO2 emission (Tons)`, Year) |>
  filter(Year==2020) |>
  arrange(desc(`CO2 emission (Tons)`))

##noteworthy countries: USA, Germany, China, South Africa, Australia, ; UK,France,Russia, Japan, India, Canada
```


```{r}
highlights <- c("USA", "China", "Russia", "Germany", "UK", "Japan")

co2_emission <- co2_emission |>
  mutate(group = if_else(Country %in% highlights, Country, "Other"),
    group = as.factor(group)
  ) |> 
  mutate(group = fct_relevel(group, "Other", after = Inf)
  ) |>
  ungroup()
#59,620 x 9
```

#But what is causing temperatures to rise?


#Let's look more closely at carbon dioxide emissions by country. 

```{r}
co2_emission |>
  filter(Year>1960 & Year<2022,
         group!="Other") |>
  ggplot() +
  aes(x=Year, y=`CO2 emission (Tons)`, group=Country ) +
  geom_line(data = co2_emission |> filter(Year>1960 & Year<2022, group == "Other"),
            color = "grey75",
            alpha = .5) +
  geom_line(aes(colour=group), linewidth=1) +
  labs(x='Year',
       y='CO2 Emission in Billion Tons',
       title = "CO2 Emission by each country",
       colour="Country") +
  scale_x_continuous(breaks = seq(1965, 2020, by = 5), expand = c(0, NA)) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) 
```



#We observe some countries contributing more than others.

```{r}
co2_emission |>
  filter(Year>=2011) |>
  mutate(Country = if_else(Country %in% c("USA", "China", "Russia", "Germany", "UK", "Japan"), Country, "Other")) |>
  group_by(Country) |>
  summarise(total_emission=sum(`CO2 emission (Tons)`)) |>
  arrange(total_emission) |>
  ggplot() +
  aes(x = "", y = total_emission, fill = factor(Country, levels = as.character(Country))) +
  geom_bar(stat = "identity", linewidth=1.5, color = "white") +
  coord_polar(theta = "y") +
  scale_fill_brewer(palette = "Oranges") +
  guides(fill = guide_legend(reverse = TRUE)) +
  theme_minimal() + 
  labs(title = "CO2 Emission Share in the last decade") +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        axis.text.x=element_blank(),
        panel.border = element_blank(),
        panel.grid=element_blank(),
        legend.position="none",
        plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(label = paste0(Country, '\n',percent(round(total_emission/sum(total_emission),2))),
                x=1.6), 
            size=4,
            position = position_stack(vjust = 0.5))
#https://r-graph-gallery.com/circle-packing.html
```


```{r}
co2_reasons <- read_excel('data/global_co2_reasons.xlsx',
                          sheet=3,
                          skip=6) 
```


#How are they emitting so much carbon dioxide? 
```{r}
co2_reasons |>
  filter(Year>1961) |>
  select(Year, Coal, Oil, Gas, `Cement emission`, Flaring, Other) |>
  pivot_longer(!"Year", names_to = "reasons", values_to = "emission") |>
  ggplot()+
  aes(x=Year, y=emission, fill=factor(reasons)) + 
  geom_area(alpha=0.75 , linewidth=1, colour="white") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(fill="Reasons",
       y="CO2 Emission",
       title="Major reasons for Carbon Emission") +
  scale_x_continuous(breaks = seq(1960, 2020, by = 5), expand = c(0, NA))
  
```


